Efficient pedigree recording for fast population genetics simulation

### 2017-12-27 forwards_paper.tex
make clear we are approximating

%% Memory reduction thanks to simplification
Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations,
with simplify interval $s = 1$ so that redundant information is removed after every generation.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
($T \sim 20N$, say).
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
This tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of the edges of a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per individual per generation.
Not all such recombinations actually change the tree topology,
so four times this gives us an upper bound on the expected number of additional edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $120 N^2$ items to
% 4N - 2 + 2N + 4N log(2N) * 4 * 2 = 6N + 32 N log(2N)
$6N + 32 N \log(2N)$;
at $N=10^4$, a factor of 3,700.

>>>>

%% Memory reduction thanks to simplification
Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations,
with simplify interval $s = 1$ so that redundant information is removed after every generation.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
($T \sim 20N$, say).
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
Ignoring the possibility of coalescent events involving more than two lineages, this tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of the edges of a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per individual per generation.
Not all such recombinations actually change the tree topology,
so four times this gives us an upper bound on the expected number of additional edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $120 N^2$ items to
% 4N - 2 + 2N + 4N log(2N) * 4 * 2 = 6N + 32 N log(2N)
$6N + 32 N \log(2N)$;
at $N=10^4$, a factor of 3,700.
